Transition path sampling based calculations of free energies of enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-04-23 tex/manuscript.tex
added little text and a figure

The reaction catalyzed by $pv$ADA is shown in  
Fig. \ref{fig:ada-reaction} which is between the adenosine molecule and a hydroxide ion
which results in the formation of inosine with ammonia as a by product. 
$pv$ADA is an essential enzyme in the purine salvage pathway for plasmodium vivax and 
inhibiting this enzyme is a promising approach to treat malaria. \cite{Madrid08JBiolChem283p35899}  
This reaction proceeds with the nucleophilic O atom (O$^{\text{hydroxyl}}$) of the hydroxide anion
attacking the electrophilic C6 atom of the adenosyl ring leading to the formation of a Meisenheimer 
complex while the N1 atom on the adenosyl ring starts to get 
protonated by the Glu229 residue of $pv$ADA. At the transition state, experimental findings 
suggest that the dissociation of the C6$-$N6 bond as well as the protonation of the N1 
atom are almost complete. \cite{Luo07JAmChemSoc129p8008} The key bond parameters are 
the distance between the approaching hydroxyl oxygen and the C6 atom (denoted as $dis(OC)$) and the 
dissociating C6$-$N6 ($dis(NC)$) bond distance. For the TPS simulaions we define the reactant state to 
have $dis(NC)$ being less than $1.7$ {\AA} and the product state to have the 
$dis(OC)$ less than $1.7$ {\AA}.

@@

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.

>>>>

$pv$ADA catalyzes the reaction between adenosine and hydroxide ion 
resulting in the formation of inosine and ammonia as shown in
Fig. \ref{fig:ada-reaction}.
$pv$ADA is an essential enzyme in the purine salvage pathway and 
inhibiting this enzyme is a promising approach to treat malaria. \cite{Madrid08JBiolChem283p35899}  
This reaction proceeds with the nucleophilic O atom (O$^{\text{hydroxyl}}$) of the hydroxide anion
attacking the electrophilic C6 atom of the adenosyl ring leading to the formation of a Meisenheimer 
complex while the N1 atom on the adenosyl ring starts to get 
protonated by the Glu229 residue of $pv$ADA. At the transition state, experimental findings 
suggest that the dissociation of the C6$-$N6 bond as well as the protonation of the N1 
atom are almost complete. \cite{Luo07JAmChemSoc129p8008} The key bond parameters are 
the distance between the approaching hydroxyl oxygen and the C6 atom (denoted as $dis(OC)$) and the 
dissociating C6$-$N6 ($dis(NC)$) bond distance. For the TPS simulaions we define the reactant state to 
have $dis(NC)$ being less than $1.7$ {\AA} and the product state to have the 
$dis(OC)$ less than $1.7$ {\AA}.

@@

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The TPS ensemble was used to analyze the reaction pathway for the S$_{\text{N}}2$
mechanism and a typical trajectory is provided here
\begin{figure}
\includegraphics[scale=0.5]{figures/mat2a-diff60.pdf}
\caption{A reactive trajectory}
\label{fig:mat2a-reactive-traj}
\end{figure}
The order parameters of the system along a typical reactive trajectory is 
shown in Fig. \ref{fig:mat2a-reactive-traj}.
